### 2015-09-24 manuscript.Rnw
get line numbers to work
small edits to intro

\usepackage{natbib}
\usepackage{lineno}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[font=sf]{caption}

@@

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the tree to be updated as relationships are elucidated or
reevaluated. The current draft of the Open Tree of Life contains 2.3 million
tips. Beyond obvious applications across the life sciences to explore research
questions in evolution, biodiversity, and conservation, the resources OTL
provides could be used in education and outreach (e.g., illustrating course
material, develop outreach activities to explore relationships among species).

@@

\texttt{rotl} gives users access to all endpoints provided by version 2 of the
OTL's API. Phylogenetic trees served by the API can be imported directly into
R's memory and are represented using the \texttt{ape} \cite{Paradis2004} tree
structure (objects of class \texttt{phylo}) or can be written to files in the
NEXUS, Newick or NeXML file formats. This allows researchers to use these trees
either directly with other R packages, or to be imported in other programs that
can read in phylogenetic tree files.

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used on the trees
returned by the API. However, resources and methods are being developed to add
branch lengths information to these topological trees (e.g., \cite{Ksepka2015},
\cite{Pennell2015}). Even without branch lengths, these trees could be useful to
illustrate relationships among species, or to map a trait on a phylogeny.

\texttt{rotl} can also be used to download trees accompanying studies, and that
are used to build the synthetic tree. All the trees from a study are available
even if they do not contribute to the synthetic tree. These trees provide a
useful resource to reproduce or expand on an already published analysis.
%% any examples available?

>>>>

% line numbers
\usepackage{lineno}
\modulolinenumbers[5]
\linenumbers

\usepackage{natbib}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[font=sf]{caption}

@@

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the Open Tree of Life contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore research questions in evolution, biodiversity, and conservation, the
resources OTL provides could be used in education and outreach (e.g.,
illustrating course material, develop outreach activities to explore
relationships among species).

@@

\texttt{rotl} gives users access to all endpoints provided by version 2 of the
OTL's APIs. Phylogenetic trees served by the API can be imported directly into
R's memory and are represented using the \texttt{ape} \cite{Paradis2004} tree
structure (objects of class \texttt{phylo}), or can be written to files in the
Newick, NEXUS \citep{Maddison1997} or NeXML \citep{Vos2012} file formats. This
allows researchers to use these trees either directly with other R packages, or
to be imported in other programs that can read in phylogenetic tree files.

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used on the trees
returned by the API. However, resources and methods are being developed to add
branch lengths information to these topological trees (e.g., \cite{Ksepka2015})
or use topological trees to identify phylogenetically equivalent species to
increase overlap between chronograms and species trait data
\cite{Pennell2015}. Without branch lengths, these trees remain useful to
illustrate relationships among species, or to map a trait on a phylogeny.

\texttt{rotl} can also be used to download trees accompanying studies that have
been submitted through the curator interface, and identify the trees that are
being used to assemble the synthetic tree. These trees provide a useful resource
to reproduce or expand on an already published analysis.

%% any examples available?
